The \project{unRAVE} catalog

### 2016-08-31 article/unrave.tex
regularization essential; start writing up all the hacks!

\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\thecannon}{\project{The~Cannon}}
\newcommand{\rave}{\project{\acronym{RAVE}}}
\newcommand{\galah}{\project{\acronym{GALAH}}}
\newcommand{\ges}{\project{Gaia-ESO}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\aspcap}{\project{\acronym{ASPCAP}}}
\newcommand{\lamost}{\project{\acronym{LAMOST}}}
\newcommand{\hipparcos}{\project{Hipparcos}}
\newcommand{\epic}{\project{K2/EPIC}}

@@

\newcommand{\teff}{T_{\mathrm{eff}}}
\newcommand{\logg}{\log g}

@@

The Milky Way is considered to be the best laboratory for understanding galaxy
formation and evolution.  This premise hinges on the ability to precisely measure 
the astrometry and chemistry for (many) individual stars, and to use those data 
to infer the structure, kinematics, and chemical enrichment of the Galaxy 
\citep[e.g.,][]{Schlaufman_2009,Deason_2011,Ness_2012,Ness_2013a,Ness_2013b,
Casey_2012,Casey_2013,Casey_2014a,Casey_2014b,Bovy_2016}.  However, these 
quantities are not known for even 1\% of stars in the Milky Way.  Stellar 
distances are famously imprecise \citep[e.g.,][]{van_Leeuwen_2007,Jofre_2015,
Madler_2016}, proper motions can be plagued by unquantified systematics from 
the first epoch observations \citep[e.g.,][]{Casey_Schlaufman_2015}, and 
stellar spectroscopists frequently report significantly different chemical 
abundance patterns \citep{Smiljanic_2014}.  The impact these issues have on our
scientific inferences cannot be understated.  They limit our understanding on a
number of sub-fields in astrophysics, including the properties of exoplanet host
stars, the formation (and destruction) of stars and clusters, as well as the 
structure of the Galactic disk, to name a few.

These formed just some of the scientific drivers for the \gaia\ space telescope.
\gaia\ is primarily an astrometric mission, providing precise positions,
parallaxes and proper motions for more than $10^9$ stars in its final data
release in 2022.  While this is a sample size about three orders of magnitude 
larger than its predecessor \hipparcos, both astrometry and chemistry are 
required to fully characterize the formation and evolution of our Galaxy.  For
this reason \gaia\ will also provide radial velocities, stellar parameters and
chemical abundances for a subset of brighter stars.  While astrometry for bright
and nearby stars will be imminently available, chemical abundances from \gaia\ 
will not be available until later releases.  Until those abundances are available,
astronomers seeking to use chemistry and kinematic measurements are reliant on 
ground-based spectroscopic surveys in order to complement the available astrometry.

@@

We briefly describe the \rave\ data in Section~\ref{sec:data}, before explaining
our method in Section~\ref{sec:method}.  In Section~\ref{sec:validation}
we outline a number of validation experiments, including: internal sanity checks,
comparisons with literature samples, and investigations to ensure our results
are consistent with expectations from astrophysics.  We discuss these results in
Section~\ref{sec:discussion}, and in Section~\ref{sec:conclusion} we provide 
concluding remarks as well as a description of how to access our results electronically.

@@

The target selection for \rave\ is based on the $I$-band apparent magnitude,
$9 < I < 12$, with a weak $J - K_s > 0.5$ cut near the disk and bulge \citep{Wonjo_2016}.  
The $I$ band was selected because it overlaps with the wavelength range that \rave\ 
operates in:  8410--8795~\AA.  This setup includes the \ion{Ca}{2} near-infrared 
triplet, and a number of atomic transitions of light-, $\alpha$-, and Fe-peak elements.
The \ion{Ca}{2} near infrared triplet lines are dominated by pressure broadening, 
which makes them strong and therefore visible even in very low S/N data or 
metal-poor stars.  This spectral region is also very well-studied.  It is one of the
key setups used for the ground-based high-resolution \ges\ survey
\citep{Gilmore_2012,Randich_2013}, and the \rave\ resolution and wavelength
coverage is comparable to the Radial Velocity Spectrometer on board the \gaia\
space telescope \citep{Recio-Blanco_2016}.

@@

Given the high-quality of the normalization performed by the \rave\ team, we chose
not to re-normalize the spectra.  Our tests demonstrated that the procedure 
outlined in \citet{Kunder_2016} is sufficient for our analysis procedure. Therefore,
there is a limited number of pre-processing steps that we conducted before our
analysis could begin.  Firstly, we calculated inverse variance arrays from the
fractional uncertainties provided, then re-sampled the flux and inverse variance
arrays onto a common rest-wavelength map for all stars.  Depending on the fibre 
used and the stellar radial velocity, the range of rest-frame wavelength values
varied for each star.  Given that fluxes were unavailable in the edge pixels for 
most stars, we excluded pixels outside of the rest wavelength range 
$8423.2\,{\rm \AA} \le \lambda \le 8777.6\,{\rm \AA}$.  This corresponds to about
30~pixels excluded on either side of the common wavelength array, leaving us with
945~pixels per spectrum for science.

@@

Throughout this work we adopt the assumptions and nomenclature first set out in
\citet{Ness_2015}, and extended in \citet{Casey_2016b}.  However, one distinction 
here with \citet{Casey_2016b} is that in this work the \emph{labelled} set and the
\emph{training} set will be the same set.  In \citet{Casey_2016b} two subsets of
the labelled set were defined: the training set and a distinct validation set.
This allowing us to train on the \emph{training set} and then test on the 
\emph{validation set}.  However in this work the labelled set is relatively 
small, about an order of magnitude smaller than what was available in 
\citet{Casey_2016b}, and therefore no validation set exists. We will use all 
labelled set stars in the training step and supplement our validation experiments 
with literature comparisons.  Thus, throughout this work we will use the terms 
\emph{labelled set} and \emph{training set} interchangeably.  Below we describe 
how we constructed the training set, before we outline the model that we adopt.

@@

We sought to construct a training set consisting of stars across the main-sequence,
sub-giant, and the red giant branch.  We required stars with precisely measured
effective temperature $\teff$, surface gravity $\logg$, and elemental abundances
of Fe, Mg, Si, Ti, Ca, Al, and Ni.  This proved to be difficult because the magnitude
range of \rave\ does not overlap substantially with high-resolution spectroscopic
surveys.  The fourth internal data release of the \ges\ survey includes 
giant and main-sequence stars, but only 142 overlap with \rave.  The most recent 
\apogee\ data release \citep{sloan_dr13} reports labels for stars on the
giant branch and (uncalibrated values for) the main-sequence, but our tests indicated
that the main-sequence labels suffered from significant systematic effects.  A flat, 
then `up-turning' main-sequence is present, and the metallicity gradient trends in 
the opposite direction with respect to $\logg$ on the main-sequence (i.e., metal-poor
stars incorrectly sit above the isochrones in a classical Hertzsprung-Russell diagram)
If we consider lower-resolution studies, there are 2,369 stars that overlap with 
\lamost\ --- of which 2,213 have positive S/N ratios in the $g$-band 
(\texttt{snrg}).  However, the labels are expectedly less precise given the lower 
resolution, no detailed abundance information is available for main-sequence stars
\footnote{Abundance information is available for \lamost\ stars from \citet{Ho_2016},
but that sample contains only giant stars}, and similar systematic trends are apparent
in the lower main-sequence.  The \galah\ survey has a good overlap with \rave, but 
results from that sample are not yet precise enough for our purposes.

@@

At the training step we fix the $K$-lists of labels for all the training set stars $n$.
At each wavelength pixel $j$, we then calculate the parameters $\vectheta_j$ and $s_j^2$
by optimizing the likelihood function:
\begin{eqnarray}\label{eq:train}
\vectheta_j,s^2_j &\leftarrow& \argmin{\vectheta,s}\left[
    \sum_{n=0}^{N-1} \frac{[y_{jn}-\vecv(\ell_n)\cdot\vectheta]^2}{\sigma^2_{jn}+s^2}
    + \sum_{n=0}^{N-1} \ln(\sigma^2_{jn}+s^2)
    \right]
  \quad .
\end{eqnarray}

In practice this is achieved by fixing $s_j^2 = 0$ to make equation~\ref{eq:train} a
convex optimization problem, then we optimize for $\vectheta_j$, before solving for $s_j^2$.
We can also adopt strict priors on $\vectheta$ such that labels can only make non-zero
contributions to the data in certain wavelength ranges.  That is to say, we can specify
that certain labels (e.g., [Al/H]) can only contribute to the model at wavelength pixels
where we know an aluminium transition exists.  This method, which we call \emph{censoring},
is described in Section~\ref{sec:censoring} and \citet{Ness_2016b}.

The \emph{test step} is where we fix the parameters $\vectheta_j,s_j^2$ at all wavelength
pixels $j$, and optimize the $K$-list of labels $\ell_m$ for each test set star $m$.  Here
the likelihood function is:
\begin{eqnarray}\label{eq:test}
  \ell_m &\leftarrow& \argmin{\ell}\left[
    \sum_{j=0}^{J-1} \frac{[y_{jm}-\vecv(\ell)\cdot\vectheta_j]^2}{\sigma_{jm}^2 + s_j^2}
    \right]
  \quad .
\end{eqnarray}

@@

\subsection{Three-label models ($\teff$, $\logg$, [Fe/H])}
\label{sec:a-simple-model}

% Given previous successes, we will first adopt a three-parameter model for giants + dwarfs.
% Polynomial thing --> expand out \vectheta
% worked but didn't show the structure in detail.
% Split into two different models with partial overlap.
% One for giants and one for dwarfs.
% The labels for stars in the training sets are shown in Figure X.

@@

\subsection{A 10-label model for detailed abundances of giant stars}
\label{sec:giant-abundance-model}
% Restricted to 3-label model, but the giant set includes abundance labels.
% The spectral range covered by \rave\ includes atomic transitions of X, Y, Z elements.
% The strongest lines are indicated in Figure XXX.
% Censoring masks

% Train

@@

% TODO: Replace ZENODO-URL with a data url
Source code for this project is available at \texttt{\giturl}.
This document was compiled from revision hash \texttt{\githash} in that repository.
Inferred labels and covariance matrices are available electronically with 
associated metadata at \texttt{\url{ZENODO-URL}} \citep{DATA_REPOSITORY}.  
Please note that it is a condition of using these results that the fifth \rave\ 
data release by \citet{Kunder_2016} must (also) be cited, as the work presented 
here would not have been possible without the tireless efforts of the entire 
\rave\ collaboration, past and present.

>>>>

\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\thecannon}{\project{The~Cannon}}
\newcommand{\rave}{\project{\acronym{RAVE}}}
\newcommand{\galah}{\project{\acronym{GALAH}}}
\newcommand{\ges}{\project{Gaia-ESO}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\aspcap}{\project{\acronym{ASPCAP}}}
\newcommand{\lamost}{\project{\acronym{LAMOST}}}
\newcommand{\hipparcos}{\project{Hipparcos}}
\newcommand{\epic}{\project{K2/EPIC}}
\newcommand{\sdss}{\project{\acronym{SDSS}}}

@@

\newcommand{\teff}{T_{\mathrm{eff}}}
\newcommand{\logg}{\log g}
\newcommand{\feh}{[\mathrm{Fe/H}]}

@@

The Milky Way is considered to be the best laboratory for understanding galaxy
formation and evolution.  This premise hinges on the ability to precisely measure 
the astrometry and chemistry for (many) individual stars, and to use those data 
to infer the structure, kinematics, and chemical enrichment of the Galaxy 
\citep[e.g.,][]{Schlaufman_2009,Deason_2011,Ness_2012,Ness_2013a,Ness_2013b,
Casey_2012,Casey_2013,Casey_2014a,Casey_2014b,Bovy_2016}.  However, these 
quantities are not known for even 1\% of stars in the Milky Way.  Stellar 
distances are famously imprecise \citep[e.g.,][]{van_Leeuwen_2007,Jofre_2015,
Madler_2016}, proper motions can be plagued by unquantified systematics from 
the first epoch observations \citep[e.g.,][]{Casey_Schlaufman_2015}, and 
stellar spectroscopists frequently report significantly different chemical 
abundance patterns for the same star \citep{Smiljanic_2014}.  The impact 
these issues have on scientific inferences cannot be understated.  Imperfect 
astrometry or chemistry limits understanding in a number of sub-fields in
astrophysics, including the properties of exoplanet host stars, the formation 
(and destruction) of stars and clusters, as well as the structure of the 
Galactic disk, to name a few.

These formed just some of the scientific drivers for the \gaia\ space telescope.
\gaia\ is primarily an astrometric mission, providing precise positions,
parallaxes and proper motions for more than $10^9$ stars in its final data
release in 2022.  While this is a sample size about three orders of magnitude 
larger than its predecessor \hipparcos, both astrometry and chemistry are 
required to fully characterize the formation and evolution of our Galaxy. 
\gaia\ will also provide radial velocities, stellar parameters and chemical 
abundances for a subset of brighter stars, but these measurements will not be 
available in the first few data releases. Until those abundances are available,
astronomers seeking to use chemistry and kinematic measurements are reliant on 
ground-based spectroscopic surveys in order to complement the available astrometry.

@@

We briefly describe the \rave\ data in Section~\ref{sec:data}, before explaining
our methods in Section~\ref{sec:method}.  In Section~\ref{sec:validation}
we outline a number of validation experiments, including: internal sanity checks,
comparisons with literature samples, and investigations to ensure our results
are consistent with expectations from astrophysics.  We discuss these results in
Section~\ref{sec:discussion}, and in Section~\ref{sec:conclusion} we provide 
concluding remarks as well as a description of how to access our results electronically.

@@

The target selection for \rave\ is based on the $I$-band apparent magnitude,
$9 < I < 12$, with a weak $J - K_s > 0.5$ cut near the disk and bulge \citep{Wonjo_2016}.  
The $I$ band was selected because it overlaps with the wavelength range that \rave\ 
operates in:  8410--8795~\AA.  This spectral region is very well-studied.  It is one 
of the key setups used for the ground-based high-resolution \ges\ survey
\citep{Gilmore_2012,Randich_2013}, and the \rave\ resolution and wavelength
coverage is comparable to the Radial Velocity Spectrometer on board the \gaia\
space telescope \citep{Recio-Blanco_2016}.  The region includes the \ion{Ca}{2} 
near infrared triplet lines, strong transitions that are dominated by pressure 
broadening, which are visible even in very low S/N data or metal-poor stars.  
Atomic transitions of light-, $\alpha$-, and Fe-peak elements are also present, 
allowing for detailed chemical abundance studies.

@@

Given the high-quality of the normalization performed by the \rave\ team, we chose
not to re-normalize the spectra.  Our tests demonstrated that the procedure 
outlined in \citet{Kunder_2016} is sufficient for our analysis procedure. Therefore,
there is a limited number of pre-processing steps that we conducted before our
analysis could begin.  First, we calculated inverse variance arrays from the
fractional uncertainties provided, and then re-sampled the flux and inverse variance
arrays onto a common rest-wavelength map for all stars.  Depending on the fibre 
used and the stellar radial velocity, the range of rest-frame wavelength values
varied for each star.  Given that fluxes were unavailable in the edge pixels for 
most stars, we excluded pixels outside of the rest wavelength range 
$8423.2\,{\rm \AA} \le \lambda \le 8777.6\,{\rm \AA}$.  This corresponds to about
30~pixels excluded on either side of the common wavelength array, leaving us with
945~pixels per spectrum for science.

@@

Throughout this work we adopt the assumptions and nomenclature first set out in
\citet{Ness_2015}, and extended in \citet{Casey_2016b}.  However, one distinction 
here with \citet{Casey_2016b} is that in this work the \emph{labelled} set and the
\emph{training} set will be the same set.  In \citet{Casey_2016b} two subsets of
the labelled set were defined: the training set and a distinct validation set.
This allowing us to train on the \emph{training set} and then test on the 
\emph{validation set}.  However in this work the labelled set is relatively 
small, about an order of magnitude smaller than what was available in 
\citet{Casey_2016b}, and therefore no validation set exists. We will use all 
labelled set stars in the training step and supplement our validation experiments 
with literature comparisons.  Thus, throughout this work we will use the terms 
\emph{labelled set} and \emph{training set} interchangeably.  Below we describe 
how we constructed the training set, before we outline the models that we adopt.

@@

We sought to construct a training set consisting of stars across the main-sequence,
sub-giant, and the red giant branch.  We required stars with precisely measured
effective temperature $\teff$, surface gravity $\logg$, and elemental abundances
of Fe, Mg, Si, Ti, Ca, Al, and Ni.  This proved to be difficult because the magnitude
range of \rave\ does not overlap substantially with high-resolution spectroscopic
surveys.  The fourth internal data release of the \ges\ survey includes 
giant and main-sequence stars, but only 142 overlap with \rave.  The thirteenth data 
release from \sdss\ \citep{sloan_dr13} includes labels for \apogee\ stars on the
giant branch and (uncalibrated values for) the main-sequence, but our tests indicated
that the main-sequence labels suffered from significant systematic effects.  A flat, 
then `up-turning' main-sequence is present, and the metallicity gradient trends in 
the opposite direction with respect to $\logg$ on the main-sequence (i.e., metal-poor
stars incorrectly sit above the isochrones in a classical Hertzsprung-Russell diagram)
If we consider lower-resolution studies as potential training sets, there are 2,369
stars that overlap with \lamost\ --- of which 2,213 have positive S/N ratios in the 
$g$-band (\texttt{snrg}).  However, the labels are expectedly less precise given the
lower resolution, no detailed abundance information is available for main-sequence 
stars\footnote{Abundance information is available for \lamost\ stars from \citet{Ho_2016},
but that sample contains only giant stars}, and similar systematic trends are apparent
in the lower main-sequence.  The \galah\ survey has a good overlap with \rave, but 
results from that sample are not yet precise enough for our purposes.

@@

At the training step we fix the $K$-lists of labels for all the training set stars $n$.
At each wavelength pixel $j$, we then calculate the parameters $\vectheta_j$ and $s_j^2$
by optimizing the penalized likelihood function:
\begin{eqnarray}\label{eq:train}
\vectheta_j,s^2_j &\leftarrow& \argmin{\vectheta,s}\left[
    \sum_{n=0}^{N-1} \frac{[y_{jn}-\vecv(\ell_n)\cdot\vectheta]^2}{\sigma^2_{jn}+s^2}
    + \sum_{n=0}^{N-1} \ln(\sigma^2_{jn}+s^2) + \Lambda{}Q(\vectheta)
    \right]
  \quad ,
\end{eqnarray}

\noindent{}where $\Lambda$ is a regularization parameter which we will heuristically set
for each model in later sections, and $Q(\vectheta)$ is a regularizing function that 
encourages $\vectheta$ values to take on zero values without breaking convexity:
\begin{eqnarray}\label{eq:regularization-function}
	Q(\vectheta) = \sum_{d=1}^{D-1} |{\theta_d}|
\end{eqnarray}

Note that the $d$ subscript here is zero-indexed; the function $Q(\vectheta)$ does not act
on the (first) $\theta_0$ coefficient, as this is a `pivot point' or offset that we do not
expect to diminish with increasing regularization (e.g., see equation 
\ref{eq:vectorizer-three-label-design-matrix}).  In practice equation \ref{eq:train} is 
optimized by fixing $s_j^2 = 0$ to make equation~\ref{eq:train} a convex optimization 
problem, then we optimize for $\vectheta_j$, before solving for $s_j^2$.

The \emph{test step} is where we fix the parameters $\vectheta_j,s_j^2$ at all wavelength
pixels $j$, and optimize the $K$-list of labels $\ell_m$ for each test set star $m$.  Here
the objective function is:
\begin{eqnarray}\label{eq:test}
  \ell_m &\leftarrow& \argmin{\ell}\left[
    \sum_{j=0}^{J-1} \frac{[y_{jm}-\vecv(\ell)\cdot\vectheta_j]^2}{\sigma_{jm}^2 + s_j^2}
    \right]
  \quad .
\end{eqnarray}

@@

\subsection{Three-label models ($\teff$, $\logg$, $\feh$)}
\label{sec:a-simple-model}

The model we have introduced is generic in the sense that there is no description for how
complex the vectorizing function $\vecv(\ell)$ should be, or how many $K$ labels are used.
Given previous successes using a second-order polynomial model 
\citep{Ness_2015,Ness_2016,Ho_2016,Casey_2016b}, we will first adopt a three-label ($\teff$, $\logg$, [Fe/H]) model for all (giant and main-sequence) stars. Thus, for our first set of
model experiments:
\begin{eqnarray}\label{eq:vectorizer-three-label}
\vecv(\ell_n) & \rightarrow & \left[1, T_{{\rm eff},n}, \logg_n, [{\rm Fe/H}]_n, T_{{\rm eff},n}^2, \logg_n\cdot{}T_{{\rm eff},n}, \feh_n\cdot{}T_{{\rm eff},n}, \logg_n^2, \feh_n\cdot\logg_n, \feh_n^2\right]
\quad ,
\end{eqnarray}

\noindent{}such that $\vecv(\ell)$ produces the design matrix:
\begin{tiny}
\begin{equation}\label{eq:vectorizer-three-label-design-matrix}
\vecv(\ell) \rightarrow \begin{bmatrix}
	1 & T_{{\rm eff},0} & \logg_{0} & [{\rm Fe/H}]_{0} & T_{{\rm eff},0}^2 & \logg_{0}\cdot{}T_{{\rm eff},0} & [{\rm Fe/H}]_{0}\cdot{}T_{{\rm eff},0} & \logg_{0}^2 & [{\rm Fe/H}]_{0}\cdot{}\logg_1 & [{\rm Fe/H}]_1^2 \\
	\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
	1 & T_{{\rm eff},n} & \logg_{n} & [{\rm Fe/H}]_{n} & T_{{\rm eff},n}^2 & \logg_{n}\cdot{}T_{{\rm eff},n} & [{\rm Fe/H}]_{n}\cdot{}T_{{\rm eff},n} & \logg_{n}^2 & [{\rm Fe/H}]_{n}\cdot{}\logg_n & [{\rm Fe/H}]_n^2 \\
	\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots \\
	1 & T_{{\rm eff},N-1} & \logg_{N-1} & [{\rm Fe/H}]_{N-1} & T_{{\rm eff},N-1}^2 & \logg_{N-1}\cdot{}T_{{\rm eff},N-1} & [{\rm Fe/H}]_{N-1}\cdot{}T_{{\rm eff},N-1} & \logg_{N-1}^2 & [{\rm Fe/H}]_{N-1}\cdot{}\logg_{N-1} & [{\rm Fe/H}]_{N-1}^2
\end{bmatrix}\quad .
\end{equation}
\end{tiny}

\subsubsection{A three-label model for main-sequence and giant stars}\label{sec:a-simple-joint-model}

\subsubsection{A three-label model for main-sequence stars}\label{sec:a-simple-model-ms-stars}

% worked but didn't show the structure in detail.
% Split into two different models with partial overlap.
% One for giants and one for dwarfs.
% The labels for stars in the training sets are shown in Figure X.

@@

\subsection{A 10-label model for detailed abundances of giant stars}
\label{sec:giant-abundance-model}

The giant stars in our training set have stellar parameters and up to 15 
individual abundances from the \apogee\ \aspcap.  Only a subset of these
elements have atomic transitions in the \rave\ wavelength region.

% The giant training set has many abundances.
% Which elements?
% Tests involving censoring? --> Just indicates that many of these lines are blended.
% So we will use regularization instead of censoring

% Set hyper-parameters by LOO-CV

% Trained this and rain it on all RAVE spectra.

\subsection{Joining the models}\label{sec:joining-the-models}

@@

% TODO: Replace ZENODO-URL with a data url
\noindent{}Source code for this project is available at \texttt{\giturl}.
This document was compiled from revision hash \texttt{\githash} in that repository.
Inferred labels and covariance matrices are available electronically with 
associated metadata at \texttt{\url{ZENODO-URL}} \citep{DATA_REPOSITORY}.  
Please note that it is a condition of using these results that the fifth \rave\ 
data release by \citet{Kunder_2016} must (also) be cited, as the work presented 
here would not have been possible without the tireless efforts of the entire 
\rave\ collaboration, past and present.
